### 2022-07-08 introduction/introduction.tex
geoff chapter 1

Many years later I started studying various formal methods in computer science.
There I learned the value of formal, mathematical language and finally
understood him. Not only does using a rigorous mathematical language avoid the
ambiguity of the natural one, it allows us to use the formal rules of reasoning
to make conclusions from initial arguments in a trustworthy manner.

@@

Since its introduction, resolution was improved in many ways. Strategies and
heuristics~\cite{lw-65-sos} to curb the explosion of the search space were
introduced, as well as complete, but less explosive variants of the calculus
\cite{cc-73-resolution-book}. Despite this progress, reasoning about equality of
objects was still hard for these methods. This changed with introduction of
superposition~\cite{bg-94-superposition} in the early 1990s.

@@

Proof assistants were essential in proving very complex mathematical theorems in a
trustworthy manner. For example, the four-color theorem \cite{gg-2008-four-color}
and the Kepler conjecture \cite{th-2015-kepler} were proven inside proof assistants,
ending the need to trust hundreds of pages of mathematical proofs. Proof
assistants are also used to show correctness of complex software such as
compilers \cite{xl-09-compcert} or operating systems kernels \cite{gk-09-sel4}.
This verified software can be used in safety-critical scenarios such as nuclear
plants or autonomous vehicles.

@@

The approach proved useful: on some benchmarks, around 77\% of the proof goals can be
discharged automatically using Sledgehammer \cite{bgkku-16-larning-fact-selector}. However, there is
clearly some untapped potential in the approach as it communicates with
automatic theorem provers through a translation. To bridge this translation gap,
attempts were made to use higher-order automatic theorem provers instead of
first-order ones \cite{ns-13-leo2sh}. This proved unfruitful as state-of-the-art
higher-order theorem provers of the time were primarily designed for small and
tricky problems. On the other hand, problems coming from hammers consist of
mostly first-order formulas, are rather large, and mostly require simple
higher-order proof steps. An example of a problem that is seemingly simple, but whose translation to first-order logic
is out of reach for even the best of first-order provers is \cite{bbbntvw-2x-mechanical-mathematicians}:

$$ \big(\sum_{i=1}^{n} i^2 + 2i + 1\big) \ieq \big(\sum_{i=1}^{n} i^2\big) +
\big(\sum_{i=1}^{n} 2i\big) + \big(\sum_{i=1}^{n} 1\big) $$ \noindent One of the
reasons why this problem is hard is that higher-order proof assistants rely on
unnamed functions, called \emph{$\lambda$-abstractions}, to represent
variable-binding mathematical operators such as summations, integrals and
limits. More precisely, the left-hand side of the above equation is usually
represented as $\cst{sum}(1, n, \lambda i. \, i\cst{*}i + 2\cst{*}i + 1)$. When
translated to first-order logic, $\lambda$-abstractions, which are ubiquitous as
all-purpose binders in higher-order logic, are obfuscated to the point that all
but the simplest  reasoning about them are out of reach for first-order provers.
However, for higher-order provers the crux of the proof is matching the axiom
$\iforall i j f g.\, \cst{sum}(i, j, \lambda x. \, f \, x \, \cst{+} \, g\,x) \ieq
(\cst{sum}(i, j, \lambda x. \, f \, x ) \, \cst{+} \, \cst{sum}(i, j, \lambda x. \, g
\, x ) )$ against the goal. This operation is relatively simple and, in general, much more efficient
than first-order reasoning about translations.

@@

\paragraph{Our Approach} To improve the automatic reasoning for higher-order
logic, we decided not to start from a blank slate and build a new theorem prover.
Rather, we start from a position of strength---from the state-of-the-art
first-order superposition theorem prover E~\cite{scv-19-e23}---and extend it to higher-order logic. In this way, the points \ref{it:fo} and
\ref{it:size} from the above wishlist are already fulfilled, as E is very
efficient and deals well with large problems.

We do not extend E to higher-order logic in one atomic step. Instead, we extend
it by adding support for features of higher-order logic one by one,
which gradually increase its reasoning capabilities. With this gradual
approach we make sure that after every extension, E's
efficiency on first-order problems did not decrease. Thus, while trying
to fulfill the point \ref{it:ho} from the wishlist we make sure that points
\ref{it:fo} and \ref{it:size} are not affected.

@@

In higher-order logic we can make statements about statements. More precisely,
it treats Boolean terms (formulas) as first-class citizens, enabling us to
model some concepts in a more natural manner. For example, we can naturally
represent statements such as ``what Jasmin says is true'' as $\iforall x. \,
\cst{says}(\cst{jasmin}, x) \iimplies x$. The last stop is to support Boolean terms
as first-class citizens.
\pagebreak[2]

@@

  Together with Alexander Bentkamp and other colleagues we developed
  a complete superposition calculus for the second phase described above
  \cite{bbtvw-21-sup-lam}. Due to the complexity of the calculus and many design
  decisions that had to be evaluated in a short amount of time we decided to
  implement the calculus in Zipperposition \cite{sc-15-simon-phd,sc-supind-17}. This prover is less efficient than
  E, but it is implemented in OCaml and it was designed to make prototyping
  ideas and experimenting with calculi much easier. In 2019 we entered the
  higher-order division of CASC theorem proving competition~\cite{gs-19-casc27} with Zipperposition. 
  It finished third, one percent point behind Leo-III
  \cite{sb-21-leo3}  and 12 percent points behind  Satallax
  \cite{cb-12-satallax}.

@@

  Inspired by this success, we implemented the complete superposition
  calculus for higher-order logic with Booleans (i.e., the third phase), that we developed with Bentkamp and other
  colleagues \cite{bbtv-21-full-ho-sup}. Developing this calculus helped us understand in which
  ways our pragmatic extension was too weak to prove hard theorems of
  higher-order logic. We further fine-tuned the heuristics and implemented new ways
  to tame the search space explosion. In 2021 we released Zipperposition 2.1
  which implemented these new features. It again won
  CASC, outperforming Zipperposition 2. It was 16 percent points ahead of the
  runner-up, Vampire 4.5 \cite{lkav-13-vampire}.

  Building on all the experience we gathered while working on Zipperposition,
  we further extended Ehoh to full higher-order logic, obtaining a prover we call
  \newehoh{}. We again modified the term representation, implemented new unification and
  matching algorithms, extended indexing data structures, and tweaked
  heuristics. Even though this extension has a larger scope than the original
  extension from E to Ehoh, the strong basis that we had built earlier made it
  manageable.
  \pagebreak[2]

  A common thread for the extensions of both E and Zipperposition is that we
  heavily relied on the availability of established higher-order provers and the
  TPTP benchmark set. Whenever we noticed that there were problems E and
  Zipperposition could not solve, but the competition could, we looked for an
  explanation for this issue. This explanation would usually be in the form of
  missing inference or heuristic. Then we would try to find more problems that
  have a similar pattern and see if the developed technique is useful on them.
  If so, we would make it part of the portfolio of approaches tried on every
  problem. The ideas for our unification procedure (Chapter \ref{ch:unif}) and
  the techniques described in Chapters \ref{ch:bools} and \ref{ch:ho-techniques}
  were developed in this way.

@@

Various researchers also experimented with the idea of extending a first-order
prover to higher-order logic. Beeson \cite{mb-04-lam-logic} implemented second-order
unification and added $\lambda$-terms to one of the best superposition
provers at the time, Otter~\cite{mcc-03-otter}. Cruanes extended Zipperposition
\cite{sc-15-simon-phd,sc-supind-17} with arithmetic, induction, and rudimentary support for
higher-order logic. Bhayat and Reger implemented a complete superposition
calculus  for higher-order logic, based on combinators  in Vampire
\cite{br-20-full-sup-w-combs}. Authors of SMT (satisfiability modulo theories) solvers followed a similar path:
they extended cvc4~\cite{cbetal-11-cvc4} and veriT~\cite{bodf-09-veriT} to
support higher-order logic \cite{brotb-19-ho-smt}.

>>>>

Many years later I started studying various formal methods in computer science.
There I learned the value of formal, mathematical language and finally
understood it. Not only does using a rigorous mathematical language avoid the
ambiguity of the natural one, it allows us to use the formal rules of reasoning
to make conclusions from initial premises in a trustworthy manner.

@@

Since its introduction, resolution was improved in many ways. Strategies and
heuristics~\cite{lw-65-sos} to curb the explosion of the search space were
introduced, as well as complete, but less explosive variants of the calculus
\cite{cc-73-resolution-book}. Despite this progress, reasoning about equality of
objects was still hard for resolution. Even though some improvements, like
paramodulation calculus \cite{nr-01-paramodulation}, improved this situation,
reasoning with equality remained impractical. This changed with introduction of
superposition~\cite{bg-94-superposition} in the early 1990s.

@@

Proof assistants were essential in proving very complex mathematical theorems in a
trustworthy manner. For example, the four-color theorem \cite{gg-2008-four-color}
and the Kepler conjecture \cite{th-2015-kepler} were proven inside proof assistants,
ending the need to trust hundreds of pages of mathematical proofs. Proof
assistants are also used to show correctness of complex software such as
compilers \cite{xl-09-compcert} and operating systems kernels \cite{gk-09-sel4}.
This verified software can be used in safety-critical scenarios such as nuclear
plants and autonomous vehicles.

@@

The approach proved useful: on some benchmarks, around 77\% of the proof goals can be
discharged automatically using Sledgehammer \cite{bgkku-16-larning-fact-selector}. However, there is
clearly some untapped potential in the approach as it communicates with
automatic theorem provers through a translation. To bridge this translation gap,
an attempt was made to use higher-order automatic theorem provers instead of
first-order ones \cite{ns-13-leo2sh}. This proved unfruitful as state-of-the-art
higher-order theorem provers at the time of this attempt were primarily designed for small
problems requiring complex higher-order reasoning. On the other hand, problems coming from hammers consist of
mostly first-order formulas, are rather large, and mostly require simple
higher-order proof steps. An example of a problem that is seemingly simple, but whose translation to first-order logic
is out of reach for even the best of first-order provers is \cite{bbbntvw-2x-mechanical-mathematicians}:

$$ \big(\sum_{i=1}^{n} i^2 + 2i + 1\big) \ieq \big(\sum_{i=1}^{n} i^2\big) +
\big(\sum_{i=1}^{n} 2i\big) + \big(\sum_{i=1}^{n} 1\big) $$ \noindent One of the
reasons why this problem is hard is that higher-order proof assistants rely on
unnamed functions, called \emph{$\lambda$-abstractions}, to represent
variable-binding mathematical operators such as summations, integrals and
limits. More precisely, the left-hand side of the above equation is usually
represented as $\cst{sum}(1, n, \lambda i. \, i\cst{*}i + 2\cst{*}i + 1)$. When
translated to first-order logic, $\lambda$-abstractions, which are ubiquitous as
all-purpose binders in higher-order logic, are obfuscated to the point that all
but the simplest  reasoning about them is out of reach for first-order provers.
However, for higher-order provers the crux of the proof is matching the axiom
$\iforall i j f g.\, \cst{sum}(i, j, \lambda x. \, f \, x \, \cst{+} \, g\,x) \ieq
(\cst{sum}(i, j, \lambda x. \, f \, x ) \, \cst{+} \, \cst{sum}(i, j, \lambda x. \, g
\, x ) )$ against the goal. This operation is relatively simple and, in general, much more efficient
than first-order reasoning about translations.

@@

\paragraph{Our Approach} To improve the automatic reasoning for higher-order
logic, my colleagues and me decided not to start from a blank slate and build a new theorem prover.
Rather, we start from a position of strength---from the state-of-the-art
first-order superposition theorem prover E~\cite{scv-19-e23}---and extend it to higher-order logic. In this way, the points \ref{it:fo} and
\ref{it:size} from the above wishlist are already fulfilled, as E is very
efficient and deals well with large problems.

We did not extend E to higher-order logic in one atomic step. Instead, we extended
it by adding support for features of higher-order logic one by one,
which gradually increase its reasoning capabilities. With this gradual
approach we made sure that after every extension, E's
efficiency on first-order problems did not decrease. Thus, while trying
to fulfill the point \ref{it:ho} from the wishlist we made sure that points
\ref{it:fo} and \ref{it:size} were not affected.

@@

In higher-order logic we can make statements about statements. More precisely,
it treats Boolean terms (formulas) as first-class citizens, enabling us to
model some concepts in a more natural manner. For example, statements such as
``what Jasmin says is true'' can be represented naturally as $\iforall x. \,
\cst{says}(\cst{jasmin}, x) \iimplies x$. The last stop is to support Boolean
terms as first-class citizens.

@@

  Alexander Bentkamp with the help of other colleagues and me developed
  a complete superposition calculus for the second phase described above
  \cite{bbtvw-21-sup-lam}. Due to the complexity of the calculus and many design
  decisions that had to be evaluated in a short amount of time we decided to
  implement the calculus in Zipperposition \cite{sc-15-simon-phd,sc-supind-17}. This prover is less efficient than
  E, but it is implemented in OCaml and it was designed to make prototyping
  ideas and experimenting with calculi much easier. In 2019 we entered the
  higher-order division of CASC theorem proving competition~\cite{gs-19-casc27} with Zipperposition. 
  It finished third, one percent point behind Leo-III
  \cite{sb-21-leo3}  and 12 percent points behind  Satallax
  \cite{cb-12-satallax}.

@@

  Inspired by this success, we implemented the complete superposition calculus
  for higher-order logic with Booleans (i.e., the third phase), that was
  developed by Bentkamp with the help of other colleagues and me
  \cite{bbtv-21-full-ho-sup}. Developing this calculus helped us understand in
  which ways our pragmatic extension was too weak to prove hard theorems of
  higher-order logic. We further fine-tuned the heuristics and implemented new
  ways to tame the search space explosion. In 2021 we released Zipperposition
  2.1 which implemented these new features. It again won CASC, outperforming
  Zipperposition 2. It was 16 percent points ahead of the runner-up, Vampire 4.5
  \cite{lkav-13-vampire}.

  Building on all the experience we gathered while working on Zipperposition,
  we further extended Ehoh to full higher-order logic, obtaining a prover called
  \newehoh{}. We again modified the term representation, implemented new unification and
  matching algorithms, extended indexing data structures, and tweaked
  heuristics. Even though this extension has a larger scope than the original
  extension from E to Ehoh, the strong basis that we had built earlier made it
  manageable.
  \pagebreak[2]

  A common thread for the extensions of both E and Zipperposition is that we
  heavily relied on the availability of established higher-order provers and the
  TPTP benchmark set. Whenever we noticed that there were problems E and
  Zipperposition could not solve, but other competitive provers could, we looked for an
  explanation for this issue. This explanation would usually be in the form of
  missing inference or heuristic. Then we would try to find more problems that
  have a similar pattern and see if the developed technique is useful on them.
  If so, we would make it part of the portfolio of approaches tried on every
  problem. The ideas for our unification procedure (Chapter \ref{ch:unif}) and
  the techniques described in Chapters \ref{ch:bools} and \ref{ch:ho-techniques}
  were developed in this way.

@@

Various researchers also experimented with the idea of extending a first-order
prover to higher-order logic. Beeson \cite{mb-04-lam-logic} implemented second-order
unification and added $\lambda$-terms to one of the best first-order
provers at the time, Otter~\cite{mcc-03-otter}. Cruanes extended Zipperposition
\cite{sc-15-simon-phd,sc-supind-17} with arithmetic, induction, and rudimentary support for
higher-order logic. Bhayat and Reger implemented a complete superposition
calculus  for higher-order logic, based on combinators  in Vampire
\cite{br-20-full-sup-w-combs}. Authors of SMT (satisfiability modulo theories) solvers followed a similar path:
they extended cvc4~\cite{cbetal-11-cvc4} and veriT~\cite{bodf-09-veriT} to
support higher-order logic \cite{brotb-19-ho-smt}.

### 2022-07-08 summary/summary.tex
geoff chapter 1

\looseness=-1
In the last decades, proof assistants were immeasurably useful in formally
proving validity of hard mathematical theories and correctness of
safety-critical software. Using these tools one can formally describe a problem
and produce machine-checkable proofs for statements about it. To make the
checking phase efficient and trustworthy, proof steps need to be of fine
granularity. As a consequence, seemingly simple statements must be proven in
minute detail, making the use of proof assistants very tedious.

>>>>

\looseness=-1
In the last decades, proof assistants have been immeasurably useful in formally
proving validity of hard mathematical theories and correctness of
safety-critical software. Using these tools one can formally describe a problem
and produce machine-checkable proofs for statements about it. To make the
checking phase efficient and trustworthy, proof steps need to be of fine
granularity. As a consequence, seemingly simple statements must be proven in
minute detail, making the use of proof assistants very tedious.
